Within chert: a multi-technique mineral and geochemical approach to the study of chert of southwestern Iberia

### 2025-04-25 analysis/paper/paper.qmd
Mac work session. Improved graphs and colours

si_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = SiO2, fill = `Formation/Type`)) +
  geom_boxplot(color = "#5C5C5C") +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_sio2, aes(x = `Formation/Type`, y = 70, label = n), size = 3) +
  coord_cartesian(ylim = c(70, 100)) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#F1B6DA", "UJE" = "#8E95C8", "MJE" = "#BA68C8", "CPT" = "#4EB3D3", "UCA" = "#93C54E", "T2-5" = "#FEE391","T6" = "#9CCFD1", "T7" = "#B1C7AB", "T8" = "#D7617A")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

ca_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = CaO, fill = `Formation/Type`)) +
  geom_boxplot(color = "#5C5C5C") +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_cao, aes(x = `Formation/Type`, y = -0.3, label = n), size = 3) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#F1B6DA", "UJE" = "#8E95C8", "MJE" = "#BA68C8", "CPT" = "#4EB3D3", "UCA" = "#93C54E", "T2-5" = "#FEE391","T6" = "#9CCFD1", "T7" = "#B1C7AB", "T8" = "#D7617A")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

ti_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = TiO2, fill = `Formation/Type`)) +
  geom_boxplot(color = "#5C5C5C") +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_tio2, aes(x = `Formation/Type`, y = -0.02, label = n), size = 3) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#F1B6DA", "UJE" = "#8E95C8", "MJE" = "#BA68C8", "CPT" = "#4EB3D3", "UCA" = "#93C54E", "T2-5" = "#FEE391","T6" = "#9CCFD1", "T7" = "#B1C7AB", "T8" = "#D7617A")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

fe_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = Fe2O3, fill = `Formation/Type`)) +
  geom_boxplot(color = "#5C5C5C") +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_fe2o3, aes(x = `Formation/Type`, y = 0, label = n), size = 3) +
  coord_cartesian(ylim = c(0, 0.75)) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#F1B6DA", "UJE" = "#8E95C8", "MJE" = "#BA68C8", "CPT" = "#4EB3D3", "UCA" = "#93C54E", "T2-5" = "#FEE391","T6" = "#9CCFD1", "T7" = "#B1C7AB", "T8" = "#D7617A")) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

## b) PCA of geological cherts: local Lower Jurassic, Central Portugal and Cadiz
pca_db_nl <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, K2O, TiO2, Fe2O3, S) |> # Removing CaO which is high NA in CPT
  filter(`Formation/Type` %in% c("LJW", "CPT", "UCA"))
# Convert "< LOD" values to NA
pca_db_nl[pca_db_nl == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_nl$`Formation/Type` <- as.factor(pca_db_nl$`Formation/Type`)  # Ensure it's a factor
group_var_nl <- pca_db_nl$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_nl <- pca_db_nl %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_nl <- complete.cases(df_numeric_nl)  # Identify complete rows
df_numeric_nl <- df_numeric_nl[complete_cases_nl, ]  # Remove NA rows
group_var_nl <- group_var_nl[complete_cases_nl]  # Filter group_var accordingly
# Scale values
df_scaled_nl <- scale(df_numeric_nl)
# Ensure row names are properly set
rownames(df_scaled_nl) <- make.names(group_var_nl, unique = TRUE)
# Perform PCA
pca_result_nl <- PCA(df_scaled_nl, graph = FALSE)
# Summary of PCA
summary(pca_result_nl)
fviz_eig(pca_result_nl, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_nl <- fviz_pca_biplot(pca_result_nl, 
                label = "var", 
                habillage = group_var_nl,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.80,
                palette = c("LJW" = "#EDA92B", "CPT" = "#4EB3D3", "UCA" = "#93C54E"),
                col.var = "#757575") +
  scale_shape_manual(values=c(17, 16, 15)) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_nl <- data.frame(pca_result_nl$ind$coord, Group = group_var_nl)
# PCA plot with groups
scatter_nl <- ggplot(df_pca_nl, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")

## c) PCA archaeological T6 compared to local Lower Jurassic and Central Portugal
pca_db_comp <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, K2O, TiO2, Fe2O3, S) |> # Removing CaO which is high NA in CPT
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "TYPE 1" ~ "T1",
    `Formation/Type` == "TYPE 2" | `Formation/Type` == "TYPE 2B" | `Formation/Type` == "TYPE 2R" ~ "T2",
    `Formation/Type` == "TYPE 3" ~ "T3",
    `Formation/Type` == "TYPE 4" ~ "T4",
    `Formation/Type` == "TYPE 5" ~ "T5",
    `Formation/Type` == "TYPE 6" | `Formation/Type` == "TYPE 6LLB" | `Formation/Type` == "TYPE 6E" ~ "T6",
    `Formation/Type` == "TYPE 7" ~ "T7",
    `Formation/Type` == "TYPE 8" ~ "T8",
    `Formation/Type` == "TYPE 9" ~ "T9",
    `Formation/Type` == "TYPE 10" ~ "T10",
    `Formation/Type` == "TYPE 11" ~ "T11",
    TRUE ~ `Formation/Type`
  )) |> 
  filter(`Formation/Type` %in% c("LJW","CPT", "T6"))
# Convert "< LOD" values to NA
pca_db_comp[pca_db_comp == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_comp$`Formation/Type` <- as.factor(pca_db_comp$`Formation/Type`)  # Ensure it's a factor
group_var_comp <- pca_db_comp$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_comp <- pca_db_comp %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_comp <- complete.cases(df_numeric_comp)  # Identify complete rows
df_numeric_comp <- df_numeric_comp[complete_cases_comp, ]  # Remove NA rows
group_var_comp <- group_var_comp[complete_cases_comp]  # Filter group_var accordingly
# Scale values
df_scaled_comp <- scale(df_numeric_comp)
# Ensure row names are properly set
rownames(df_scaled_comp) <- make.names(group_var_comp, unique = TRUE)
# Perform PCA
pca_result_comp <- PCA(df_scaled_comp, graph = FALSE)
# Summary of PCA
summary(pca_result_comp)
fviz_eig(pca_result_comp, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_comp <- fviz_pca_biplot(pca_result_comp, 
                label = "var", 
                habillage = group_var_comp,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.80,
                palette = c("LJW" = "#EDA92B", "T6" = "#7FCDBB", "CPT" = "#4EB3D3"),
                col.var = "#757575") +
  scale_shape_manual(values=c(17, 16, 15)) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_comp <- data.frame(pca_result_comp$ind$coord, Group = group_var_comp)
# PCA plot with groups
scatter_comp <- ggplot(df_pca_comp, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")

>>>>

si_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = SiO2, fill = `Formation/Type`)) +
  geom_boxplot(aes(color = `Formation/Type`), alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_sio2, aes(x = `Formation/Type`, y = 70, label = n), size = 3) +
  coord_cartesian(ylim = c(70, 100)) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#D7617A", "UJE" = "#5C6BC0", "MJE" = "#BA68C8", "CPT" = "#114D94", "UCA" = "#93C54E", "T2-5" = "#EF6C00","T6" = "#4EB3D3", "T7" = "#00796B", "T8" = "#777")) +
  scale_color_manual(values = c("LJW" = "#EDA92B", "UJW" = "#D7617A", "UJE" = "#5C6BC0", "MJE" = "#BA68C8", "CPT" = "#114D94", "UCA" = "#93C54E", "T2-5" = "#EF6C00","T6" = "#4EB3D3", "T7" = "#00796B", "T8" = "#777"), guide = "none") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

ca_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = CaO, fill = `Formation/Type`)) +
  geom_boxplot(aes(color = `Formation/Type`), alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_cao, aes(x = `Formation/Type`, y = -0.3, label = n), size = 3) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#D7617A", "UJE" = "#5C6BC0", "MJE" = "#BA68C8", "CPT" = "#114D94", "UCA" = "#93C54E", "T2-5" = "#EF6C00","T6" = "#4EB3D3", "T7" = "#00796B", "T8" = "#777")) +
  scale_color_manual(values = c("LJW" = "#EDA92B", "UJW" = "#D7617A", "UJE" = "#5C6BC0", "MJE" = "#BA68C8", "CPT" = "#114D94", "UCA" = "#93C54E", "T2-5" = "#EF6C00","T6" = "#4EB3D3", "T7" = "#00796B", "T8" = "#777"), guide = "none") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

ti_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = TiO2, fill = `Formation/Type`)) +
  geom_boxplot(aes(color = `Formation/Type`), alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_tio2, aes(x = `Formation/Type`, y = -0.02, label = n), size = 3) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#D7617A", "UJE" = "#5C6BC0", "MJE" = "#BA68C8", "CPT" = "#114D94", "UCA" = "#93C54E", "T2-5" = "#EF6C00","T6" = "#4EB3D3", "T7" = "#00796B", "T8" = "#777")) +
  scale_color_manual(values = c("LJW" = "#EDA92B", "UJW" = "#D7617A", "UJE" = "#5C6BC0", "MJE" = "#BA68C8", "CPT" = "#114D94", "UCA" = "#93C54E", "T2-5" = "#EF6C00","T6" = "#4EB3D3", "T7" = "#00796B", "T8" = "#777"), guide = "none") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

fe_boxplot <- boxplots_pxrf |> 
  ggplot(aes(x = `Formation/Type`, y = Fe2O3, fill = `Formation/Type`)) +
  geom_boxplot(aes(color = `Formation/Type`), alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", shape = 18) +
  #geom_text(data = count_data_fe2o3, aes(x = `Formation/Type`, y = 0, label = n), size = 3) +
  coord_cartesian(ylim = c(0, 0.75)) +
  theme_minimal() +
  scale_fill_manual(values = c("LJW" = "#EDA92B", "UJW" = "#D7617A", "UJE" = "#5C6BC0", "MJE" = "#BA68C8", "CPT" = "#114D94", "UCA" = "#93C54E", "T2-5" = "#EF6C00","T6" = "#4EB3D3", "T7" = "#00796B", "T8" = "#777")) +
  scale_color_manual(values = c("LJW" = "#EDA92B", "UJW" = "#D7617A", "UJE" = "#5C6BC0", "MJE" = "#BA68C8", "CPT" = "#114D94", "UCA" = "#93C54E", "T2-5" = "#EF6C00","T6" = "#4EB3D3", "T7" = "#00796B", "T8" = "#777"), guide = "none") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank())

@@

## b) PCA of geological cherts: local Lower Jurassic, Central Portugal and Cadiz
pca_db_nl <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, K2O, TiO2, Fe2O3, S) |> # Removing CaO which is high NA in CPT
  filter(`Formation/Type` %in% c("LJW", "CPT", "UCA"))
# Convert "< LOD" values to NA
pca_db_nl[pca_db_nl == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_nl$`Formation/Type` <- as.factor(pca_db_nl$`Formation/Type`)  # Ensure it's a factor
group_var_nl <- pca_db_nl$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_nl <- pca_db_nl %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_nl <- complete.cases(df_numeric_nl)  # Identify complete rows
df_numeric_nl <- df_numeric_nl[complete_cases_nl, ]  # Remove NA rows
group_var_nl <- group_var_nl[complete_cases_nl]  # Filter group_var accordingly
# Scale values
df_scaled_nl <- scale(df_numeric_nl)
# Ensure row names are properly set
rownames(df_scaled_nl) <- make.names(group_var_nl, unique = TRUE)
# Perform PCA
pca_result_nl <- PCA(df_scaled_nl, graph = FALSE)
# Summary of PCA
summary(pca_result_nl)
fviz_eig(pca_result_nl, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_nl <- fviz_pca_biplot(pca_result_nl, 
                label = "var", 
                habillage = group_var_nl,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.80,
                palette = c("LJW" = "#EDA92B", "CPT" = "#114D94", "UCA" = "#93C54E"),
                col.var = "#757575") +
  scale_shape_manual(values=c(17, 16, 15)) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_nl <- data.frame(pca_result_nl$ind$coord, Group = group_var_nl)
# PCA plot with groups
scatter_nl <- ggplot(df_pca_nl, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")

## c) PCA archaeological T6 compared to local Lower Jurassic and Central Portugal
pca_db_comp <- pxrf_raw |> 
  select(`Formation/Type`, P2O5, K2O, TiO2, Fe2O3, S) |> # Removing CaO which is high NA in CPT
  mutate(`Formation/Type` = case_when(
    `Formation/Type` == "TYPE 1" ~ "T1",
    `Formation/Type` == "TYPE 2" | `Formation/Type` == "TYPE 2B" | `Formation/Type` == "TYPE 2R" ~ "T2",
    `Formation/Type` == "TYPE 3" ~ "T3",
    `Formation/Type` == "TYPE 4" ~ "T4",
    `Formation/Type` == "TYPE 5" ~ "T5",
    `Formation/Type` == "TYPE 6" | `Formation/Type` == "TYPE 6LLB" | `Formation/Type` == "TYPE 6E" ~ "T6",
    `Formation/Type` == "TYPE 7" ~ "T7",
    `Formation/Type` == "TYPE 8" ~ "T8",
    `Formation/Type` == "TYPE 9" ~ "T9",
    `Formation/Type` == "TYPE 10" ~ "T10",
    `Formation/Type` == "TYPE 11" ~ "T11",
    TRUE ~ `Formation/Type`
  )) |> 
  filter(`Formation/Type` %in% c("LJW","CPT", "T6"))
# Convert "< LOD" values to NA
pca_db_comp[pca_db_comp == "< LOD"] <- NA
# Keep Group column separately before PCA
pca_db_comp$`Formation/Type` <- as.factor(pca_db_comp$`Formation/Type`)  # Ensure it's a factor
group_var_comp <- pca_db_comp$`Formation/Type`  
# Convert numeric columns correctly
df_numeric_comp <- pca_db_comp %>%
  select(-`Formation/Type`) %>%  # Remove Group for PCA
  mutate(across(everything(), as.numeric))  # Convert all columns to numeric
# Remove rows with NA (important: also filter group_var!)
complete_cases_comp <- complete.cases(df_numeric_comp)  # Identify complete rows
df_numeric_comp <- df_numeric_comp[complete_cases_comp, ]  # Remove NA rows
group_var_comp <- group_var_comp[complete_cases_comp]  # Filter group_var accordingly
# Scale values
df_scaled_comp <- scale(df_numeric_comp)
# Ensure row names are properly set
rownames(df_scaled_comp) <- make.names(group_var_comp, unique = TRUE)
# Perform PCA
pca_result_comp <- PCA(df_scaled_comp, graph = FALSE)
# Summary of PCA
summary(pca_result_comp)
fviz_eig(pca_result_comp, addlabels = TRUE, ylim = c(0, 100))
# PCA Biplot with Groups
biplot_comp <- fviz_pca_biplot(pca_result_comp, 
                label = "var", 
                habillage = group_var_comp,  # Use stored group column
                addEllipses = TRUE, 
                ellipse.level = 0.80,
                palette = c("LJW" = "#EDA92B", "T6" = "#4EB3D3", "CPT" = "#114D94"),
                col.var = "#757575") +
  scale_shape_manual(values=c(17, 16, 15)) +
  ggtitle(NULL)
# Add Group column back for visualization
df_pca_comp <- data.frame(pca_result_comp$ind$coord, Group = group_var_comp)
# PCA plot with groups
scatter_comp <- ggplot(df_pca_comp, aes(x = Dim.1, y = Dim.2, color = Group, shape = Group)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse(level = 0.95, aes(fill = Group), alpha = 0.2, geom = "polygon") +
  theme_minimal() +
  labs(title = "PCA of pXRF Data", x = "PC1", y = "PC2") +
  theme(legend.position = "right")
